Forage site distributions quantify collective foraging in honey bee colonies

### 2022-01-27 analysis/manuscript/paper.Rmd
changed analysis to fit models to waggle run duration

---
title: "Forage site distributions show social and private information use in honeybees "
author: "Joseph Palmer, Ash Samuelson, Elli Leadbeater and Vincent Jansen"
output:
  word_document
always_allow_html: true
bibliography: paper-refs.bib
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(showtext)
showtext_auto()

@@

@@

Environmental conditions are widely understood to influence the value of social and private information when foraging. Exploring this relationship in detail requires quantifying the extent to which foraging decisions are made using social or private information. In honeybees, foragers use the waggle dance to organise collective foraging, yet under some conditions it’s use is less effective than prioritising private information. Identifying how information is used at a collective level is challenging due to the difficulty in evaluating individual behaviours on the colony scale. Here, we outline a model to quantify the extent to which honeybees use the waggle dance by evaluating the foraging distance distributions reported on the dance floor. In confronting this model with waggle dance data from twenty hives, we demonstrate that waggle dance use depends on the local environment of a hive. Further, by combining our estimates of waggle dance use with local landscape characteristics, our methods demonstrate a means of visualising the relationship between major landscape type and waggle dance use. These methods offer a novel approach to evaluate what factors likely led to the evolution of this unique behaviour, as well as helping determine how honeybees might adapt to an increasingly anthropogenic world.

@@

Decision making by social insects typically requires the use of both social and private information (ref). However, to what degree social or private information is needed depends on circumstance (ref), and the mixture of private and social information varies both spatially and temporally. Although collective decision making using social information is an adaption to foraging in challenging environments (ref), in some cases, groups have been shown to do better when individuals prioritise information obtained through their own experiences to that provided by conspecifics [@King2007; @Rieucau2011; @Gruter2013]. Identifying the conditions which favour the use of these two types of information is important to understand how social insects make collective decisions, and determining what conditions might have led to the evolution of sociality as well as how social insects might respond to anthropogenic changes.

Honeybees provide an ideal study system in which to evaluate social information use thanks to the famous "waggle dance", which conveys the location of a profitable resource as a vector of distance (duration of the waggle run) and direction (angle of the dance) [@Frisch1993]. Resource “quality” is not encoded in the form of the dance itself, but in the number of waggle runs performed [@Frisch1993; @Boch1956; @Esch1961; @Seeley1994; @Seeley2000]. Although bees that follow dances do not specifically interpret this information on an individual level [@Hasenjager2021], the resulting over-representation of high-quality sites on the dance-floor means that they are more likely to encounter dances that advertise better forage [@Seeley1995]. This creates a feedback mechanism through which the hive processes the information collected and proportionally selects the most profitable resources [@Seeley1994A].

Despite the efficiency benefits the waggle dance reportedly provides, its contribution to colony foraging success remains unclear. In many situations, colonies that are prevented from communicating via dance behaviour achieve equal or even greater foraging success than their wild-type counterparts [@Sherman2002; @Dornhaus2004; @Price2019]. In some ecological circumstances, individuals that forgo waiting for dance information in favour of individual search, termed “scouting”, may find food more quickly, and evidence suggests that bees adopt this strategy if the benefits of dance-following drop [@Gruter2008]. Whilst individuals can be observed under experimental conditions to determine how they use different sources of information, translating this up to a colony level is challenging due to honeybees small size, extensive foraging ranges and shear number of foragers. New methods which can quantify waggle dance use at a colony level are therefore needed if we are to assess the conditions where the dance provides the most benefit and derive the conditions user which this unique behaviour evolved.

Whether an individual forager uses social information and follows a waggle dance from the dance floor or acquires private information through scouting, all foragers report resource information on the dance-floor [@Seeley1995]. Consequently, the difference in foraging behaviours between recruited individuals following dances and scouting individuals relying on private information should manifest itself in the pattern of dances reported on the dance-floor. Here, we will demonstrate that in honeybees it is possible quantify the use of social and private information through the observation of honeybees’ waggle dances.

To reveal what foraging patterns might look like under a social or personal information use strategy, we simulated honeybees foraging in a landscape using both personal and waggle dance information (Fig. 1). As it is known in the simulation which individuals forage under what strategy, we can compare the distributions of foraging distances reported on the dance floor. Figs 1c and 1d show that the shapes of the resource distance distributions under the two simulated scenarios are very different. The distribution of the first simulation is akin to that of an exponential distribution (Fig. 1 C), which is the nearest neighbour distance distribution for foragers operating in a one-dimensional environment (see methods). The distribution of the distances reported by recruits (Fig. 1 D) appears to be a Rayleigh distribution which is the nearest neighbour distribution of a two-dimensional environment (see methods).

@@

Scouts and recruits report different foraging distance distributions on the dance-floor. This means the distances reported on the dance floor can be used to quantify the relative use of private and social information. Foragers that use private information (scouts) report a distribution of distances that reflects their probability of encountering a resource in the environment (see methods), whilst foragers that use social information (recruits) report distances that reflect the distribution of the most profitable resource available (see methods). The distance distribution reported on the dance floor therefore carries information about the extent to which private and social information is used.

On the dance-floor, the distribution of distances is modified through the profitability of a resource reported. Honey bees achieve this by measuring the energetic efficiency of a foraging trip through the ratio of energetic gain to energetic cost [@Seeley1994A]. Combined with the distance distributions identified in our simulations, we can describe the distances reported on the dance-floor as a superposition of private and social information use (Fig 2, see Methods). We assume that scouts report distances as they encounter resources on their scouting forays. If resources are distributed randomly this is approximated by an exponential distribution. Recruits report the most profitable resource among the resources advertised on the dance floor and follow a Rayleigh distribution (see Methods).

This yields two models for foraging: one model based on collective decision making which uses a mixture of private and social information, comprised of the proportion of dances following an individual foraging strategy (scouts), denoted by the parameter $p$, along with the proportion of dances, $1-p$, following a waggle dance use strategy (recruit), and a model based on the sole use of private information (scouts only, $p = 1$) (Fig 2). By using maximum likelihood methods and model selection, we inferred if and, by estimating the parameter $p$, to what extent colonies use social information.

@@

To evaluate collective decision making in honey bee colonies foraging on 'natural' landscapes we analysed a dataset of 3378 waggle dance observations from 20 observation hives between April-September 2017, previously described in [@Samuelson2021]. Hives were situated at different locations in South East England (see methods, figure 3A) and were visited every two weeks for a period of 24 weeks. On each visit, two hours of continuous waggle dance data were recorded by training a camcorder onto the dance floor. The footage of the dances was decoded manually to identify reported foraging locations [@Couvillon2012; @Schurch2019]. For each site we fit both the collective and individual foraging models. We used model selection to determine which model provided the better explanation of the data, and quantify the relative use of social information through estimating the parameter $p$. We calculated goodness-of-fit using a Kolmorgorov-Smirnov (KS) test to ascertain if the model provided a plausible explanation of the data.

For 17 out of 20 study hives, a model of collective foraging provided a better explanation of the data than one of individual foraging (Fig. 3a), in 3 sites the individual model provided the best explanation of foraging patterns. Across 17 of the sites, the collective model had a good fit to the empirical foraging distances, whereas the scout model was significantly different to the data in 8 of the sites, and therefore not deemed a plausible model (Fig. 3.b). The sites shown in Figs  3c-d are representative examples showing the model fits. For the data in Fig 3c the collective model describes foraging distances extremely well and no significant difference is observed between the model and empirical foraging distances (AIC: 569, Bootstrapped KS statistic: d = 0.06, p = 0.395). In contrast the scout model and empirical foraging distances are significantly different, providing a worse explanation of the data (AIC: 622, KS statistic: d = 0.122, p < 0.01). For Fig 3d the collective model provides a slightly better fit (KS: d = 0.04, p = 0.999) than the scout model (KS: d = 0.05, p = 0.998), however, the scout model is more parsimonious (AIC: 388) than the collective model (AIC: 393).

```{r, fig.cap="Figure 3. A. Location of study hives in Southern England coloured by which model, collective (black circles) or individual (red triangles) provided the best explanation of the data as indicated by lowest AIC score. Sites STU and ZSL (C & D) are marked. Inset plot shows the location of the sites within the UK. B. Distribution of goodness of fit confidence values for each model fit to waggle dance derived foraging distances from each site. P value is derived from a bootstrapped two-sided KS test comparing the fitted model predictions to the empirical data. The red dashed line marks the significance threshold of 0.05. Values greater than 0.05 indicate no statistically significant difference is observed between the model and the data, indicating the model provides a good fit to the data. C. Model fits to waggle dance derived foraging distances collected from the ZSL hive along with histogram showing distribution. D. Model fits to waggle dance derived foraging distances collected from the STU hive along with histogram showing distribution. Model fits show the compliment cumulative frequency distribution, giving the probability of sampling a value greater than or equal to x."}
knitr::include_graphics("../results/figures/sites_model_plot.png")
```

Our results show that, whilst foraging decisions are primarily reached using social and private information, in some circumstances, colony foraging is best explained by a model of individual foraging and private information use. This suggests that in some environments, honeybees actively ignore the waggle dance and instead prioritise the use of private information. Previous empirical studies have shown that individual foragers flexibly alter their use of waggle dance information in response to unrewarding feeders [@Gruter2011], suggesting experience drives information use at the individual level. Evaluating how these individual decisions influence the collective, however, has historically been a challenge as it is effectively impossible to track an individuals foraging behaviour over a landscape.

As well as providing direct evidence that honey bees forage collectively according to a widely accepted theory [@Seeley1994; @Sherman2002], in showing that foraging patterns in three of our sites are best explained by individual foraging, our results demonstrate honey bee colonies tune waggle dance use in relation to environmental factors.

Using the $p$ parameter estimated from our model fits, we evaluated how the use of social information varies across different environments. Each site used in this study has a unique landscape comprised of different proportions of common land-use types. We first classified the different land-use types of the area surrounding each site to obtain a standardised land-use profile for the urban and agri-rural environments separately. As many of the resulting land-use types covaried with each other, we performed a Partial Least Squares (PLS) analysis to determine the combination of land-use types which explained the most variation in the proportion of scout dances.

In the Agri-rural environment a single PC explained ~60% of the variation in the proportion of scouts (beta regression: $R^2$ = 0.6, $\phi$ = 4.9, p < 0.05, Fig. 4. A). This PC correlates positively with arable land and negatively with non-agricultural unmanaged green space (NAUMGS), built-up areas and water (Fig. 4. B). This analysis allows us to visualise what aspects of a colonies local environment most influence waggle dance use. As arable land increases whilst NAUMGS and built-up areas decrease, the proportion of individuals prioritising private information increases.

```{r, fig.cap="Figure 4. A. Estimated proportion of scouts for each site in the Agri-rural landscape against the first principle component derived from a Partial Least Squares analysis of land-use type. A beta regression shows the relationship (black line) between PC1 and the proportion of scouts, with 95% CI shown by the blue shaded area. B. Loadings for PC1 showing the correlations between PC1 and each land-use type (note NAUMGS stands for Non-agricultural unmanaged green space). All points outside the shaded area significantly contribute to PC1. The colours of these land-use types are shown in the local environment of sites HOR, SWP and CAD (maps in bottom left) chosen as representative sites along the model gradient."}
knitr::include_graphics("../results/figures/lu_prop_scouts_agrirural.png")
```

A similar effect is also observed in the urban environment, where a single PC explained ~61% of the variance in the proportion of scouts (beta regression: $R^2$ = 0.61, $\phi$ = 7.4, p < 0.05, Fig. 5. A). This PC correlates positively with continuous central land, dense residential land and water, whilst correlating negatively with sparse residential and amenity grassland (Fig. 5. B). This relationship appears to be largely driven by a single site showing a model of only individual foraging as the best fit. Visualising how waggle dance use is influenced by changes in these land-use types shows that when the landscape is dominated by sparse residential land, as in site HOR, waggle dance use dominates the colonies foraging strategy. However, this reduces as sparse residential land is replaced by continuous central and dense residential lands (Fig. 5)

```{r, fig.cap="Figure 4. A. Estimated proportion of scouts for each site in the urban landscape against the first principle component derived from a Partial Least Squares analysis of land-use type. A beta regression shows the relationship (black line) between PC1 and the proportion of scouts, with 95% CI shown by the blue shaded area. B. Loadings for PC1 showing the correlations between PC1 and each land-use type. All points outside the shaded area significantly contribute to PC1. The colours of these land-use types are shown in the local environment of sites MEL, SOM and BEL (maps in bottom left) chosen as representative sites along the model gradient."}
knitr::include_graphics("../results/figures/lu_prop_scouts_urban.png")
```

The fitting of our model to waggle dance observations provides an innovative new tool to quantify the relative importance of social and private information in honeybees. Further, by identifying the land-use combinations which most influence variation in the proportion of scouts, our analysis allows us to map the environment along the major axis of honeybee information use and visualise how land-use influences the use of the waggle dance. Despite being limited to 10 sites in an urban and agri-rural environment, our results appear to agree with previous findings that the environment influences waggle dance use [@Dornhaus2006; @Price2019].

In both the agri-rural and Urban environments, land-use types which associate with resource scarcity correlate positively with a higher proportion of scouts, whilst those which correlate with resource abundance correlate with a lower proportion of scouts, indicating a greater use of the waggle dance. Arable land is typically nutritionally poor for honey bees (ref). On the contrary, non-agricultural unmanaged green space is typically nutrient rich, as are built up areas in agri-rural environments, which often have gardens with ample flowers available for pollinators (ref). The same can also be said of the urban land-use types, where continuos central and dense residential areas are also nutritionally poor (ref) compared to sparse residential and amenity grasslands. In both landscapes, water shows as significant contributor but in opposite directions, increasing waggle dance use in the agri-rural environment whilst decreasing it in the urban. This might indicate that agri-rural water ways provide good floral resources for honeybees, whereas urban water ways do not. However, the map visualisations suggests that these land-use types offer little contribution to the overall environment. More data is therefore required to evaluate this relationship further.

Overall, we have demonstrated here that the analysis of waggle dance data can make an important contribution to our understanding of social information use and provides a methodology to further evaluate how honey bees use the unique waggle dance. Recently, technological advances have emerged which enable colony metrics to be collected faster, more accurately and over greater time spans than could be gathered by hand, allowing individuals to be tracked within colonies and theories of individual behaviour to be evaluated in more depth than could have been done previously [@Wild2021]. We are not yet in a position to be able to track behaviours outside the colony in the same detail, nevertheless, by piecing together the behavioural response of individuals into a mathematical description of colony foraging, the methodology outlined here extends our ability to predict individual behaviour beyond the colony walls and in to the environment. With the advances in the decoding of the waggle dance through automated methods [@Wario2017], we face the prospect of waggle dance data becoming “big data”. Our methodology thus provides a means of analysing such large data sets to inform the debate about the importance of social information, as well as providing useful colony metrics of foraging activity.

>>>>

---
title: "Forage site distributions quantify collective foraging in honey bee colonies"
author: "Joseph Palmer, Ash Samuelson, Elli Leadbeater and Vincent Jansen"
output:
  word_document
always_allow_html: true
bibliography: paper-refs.bib
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(showtext)
showtext_auto()

@@

Alternative title:

- Waggle run durations quantify collective foraging in honey bee colonies
- Waggle run duration distributions quantify collective foraging in honey bee colonies
- Distributions of waggle dance durations quantifies collective foraging in honey bee colonies

Or something like this.

@@

Honeybee foraging is a collective behaviour that is directed through the extraordinary waggle dance communication system, whereby colony foraging effort is allocated through a series of feedback loops. Despite having been described as the most astounding example of non-primate communication, recent evidence suggests that the benefits of using dance information vary across environments and are sometimes even negligible. Yet, because we cannot easily quantify collective foraging, pinpointing which environments promote collective foraging is challenging. Here, we present a mathematical model which allows inferring the proportion of 'scout' and 'recruit' dances on a hive's dance floor and quantify collective foraging. By applying this methodology to waggle dance data from twenty hives, we show that dance information may be used less in environments where resources are sparse, potentially because colonies are required to send out more scouts to find food in these circumstances. Our methodology provides a tool to quantify waggle dance use in undisturbed hives at a much larger scale than hitherto has been possible and to identify those ecological conditions that may have driven the evolution of this remarkable collective behaviour.

Needs referencing.

@@

Within honeybee colonies, a series of simple rules that determine when and how much bees perform the celebrated waggle dance (Box 1) mean that choices between feeding sites occur at the level of the group rather than the individual. For example, because the number of dance circuits performed by a bee on returning from a food source reflects the net energetic benefits of the trip, more of the colony's workforce will be recruited to the sweeter of two equidistant sources (ref), or the closer of two equally sweet sources (ref), without requiring any individual bee to compare options. This extraordinary system is a key example of how social insect colony behaviour can take on a form extending beyond that of the individual units [@Sumpter2006].

Despite such impressive complexity, in some situations, colonies that are prevented from communicating via dances achieve equal or even greater foraging success than their wild-type counterparts [@Sherman2002; @Dornhaus2006; @Gruter2013; @Price2019]. This indicates that bees frequently ignore the spatial information in dances.

The prevalent explanation for this apparent redundancy portrays dance-based collective foraging as an evolutionary anachronism that was useful in ancient tropical environments but offers little benefit in modern temperate environments (ref). However, fossil evidence has questioned the long-held assumption that the genus Apis evolved in tropical climates (Kotthoff et al 2013). Moreover, given that dancing comes at a significant time cost (ref), this raises the issue of why all temperate *Apis mellifera* colonies continue to dance. The dual questions of what foraging selection pressures drove the evolution of this uniquely complex collective behaviour, and why it continues to exist, are thus yet to be satisfactorily answered. Empirical studies have shown that honeybee colonies use dance information flexibly in response to their environment, suggesting that bees cease to rely on dance information if it provides little benefit [@Price2019]. Thus, identifying those environments whereby modern honeybee colonies forage collectively rather than through individual search is a key step towards understanding dance evolution (ref).

---
**Box 1**

[figure of waggle dance, maybe have text in like a box or something? Or just the text]

Honeybees provide an ideal study system in which to evaluate collective foraging thanks to the famous “waggle dance”, which conveys information about location of a resource as a vector of distance (duration of the waggle run) and direction (angle of the dance) (von Frisch 1993). Through the observation and decoding of the waggle dance, a colony's dance floor provides a unique opportunity to eavesdrop on the communication and decision making leading to collective foraging decisions. Resource “quality” is conveyed through the number of waggle runs performed [@Frisch1993; @Boch1956; @Esch1961; @Seeley1994; @Seeley2000]). Although bees that follow dances do not specifically interpret this information on an individual level [@Hasenjager2021], the resulting over-representation of high-quality sites on the dance-floor means that they are more likely to encounter dances that advertise better forage [@Seeley1995], and provides the colony with a mechanism to select the most profitable resources in their environment [@Seeley1994A].
---

Overall, evidence that colony-level collective foraging offers any benefit over individual search has proved surprisingly hard to come by, mainly because this requires to distinguish scouts from recruits. Scouts are foragers that have located a resource through individual search behaviour, whilst recruits are foragers that have found a forage site by using the information provided in the dance of another bee. Scouting and recruiting are not fixed behavioural categories, because individual bees can engage in both over the course of their foraging lifetime, and foragers can dance on return from any successful trip irrespective of whether they were recruited to the forage site or found it individually. So far, scouts and recruits are distinguished by means of the foragers' individual search history, which requires following individual bees and is time and labour intensive.

Here, we will use a different approach and develop a model that describes the proportion of foraging trips in which bees act as scouts, and as recruits, within a colony, based on the distribution of foraging distances reported by waggle dancing, encoded as the duration of the run phase of the dance [@Frisch1993]. We will use this model to infer the proportion of scouts and recruits on the dance floor from dance data, and to demonstrate the principle, apply this methodology to dance data from a collection of hives in different environments.

To reveal what the distribution of waggle run durations reported on the dance floor look like, and to demonstrate that dances from recruits and scouts lead to different distributions, we simulated honeybees foraging in a landscape where resource patches were randomly placed in the environment. Foragers could locate these under two different strategies: either acting as a scout and locating resources themselves, or following a recruit strategy and locating resources by following a random dance from the dance floor (Fig. 1). As it is known in the simulation which individuals forage under what strategy, we can compare the distributions of foraging distances reported on the dance floor. Figs 1c and 1d show that the shapes of the resource distance distributions for the two types of foraging trips are very different. The distance distribution of the scouts is akin to that of an exponential distribution (Fig. 1 C), which is the nearest neighbour distance distribution for foragers operating in a one-dimensional environment (see methods). The distribution of the distances reported by recruits (Fig. 1 D) is a Rayleigh distribution which is the nearest-neighbour distribution of a two-dimensional environment [@Pyke1978] (Pyke 1978) (see Methods).

@@

Scouts and recruits create different foraging distance distributions on the dance-floor. These different distance distributions can be used to quantify the relative importance of collective foraging. Foragers that scout report a distribution of distances that reflects their probability of encountering a resource in the environment (see methods), whilst foragers that recruits report distances that reflect the distribution of the most profitable resource available (see methods). On the dance-floor, the distribution of distances is modified through the number of runs performed for a resource, which depends on the profitability of a resource, which honeybees achieve by measuring the energetic efficiency of a foraging trip through the ratio of energetic gain to energetic cost [@Seeley1994A]. By combining this profitability with the distributions identified in our simulations, we develop a method to describe the distribution of waggle runs reported on the dance-floors of real honeybee colonies as a superposition of scout and recruit distributions (Fig 2, see Methods).

This method intrinsically captures honeybee foraging as collective decision making as the foraging sites represented on the dance floor derive from a mixture of individual search and waggle dance information, captured in the proportions p and 1-p, respectively. Consequently, by setting p to 1, we can also derive a model based on the sole use of individual search (scouts only) (Fig 2). By fitting these two models to waggle run duration distributions from real honeybee colonies we are thus able, using model selection [@BurnhamAnderson?](Burnham and Anderson), to infer if and, by estimating the parameter p, to what extent honeybee colonies use social information when foraging.

@@

To evaluate the use of waggle dance information and individual search in honeybee colonies foraging in 'natural' landscapes, we analysed an existing dataset of 3378 waggle dance observations from 20 observation hives, recorded between April-September 2017, (previously described in [@Samuelson2021]). Hives were situated at different locations in South East England (see methods, figure 3A) and were visited every two weeks for a period of 24 weeks. On each visit, two hours of continuous waggle dance data was recorded by training a camcorder onto the dance floor. The footage of the dances was decoded manually ([@Couvillon2012; @Schurch2019] to extract waggle run durations. For each site we fit both the collective and individual models and used model selection to determine which provided the better explanation of the data, and (if the collective model provided a better fit) to quantify the relative use of social information through estimating the parameter p. We calculated the goodness-of-fit using a Kolmorgorov-Smirnov (KS) test to ascertain if the model provided a plausible explanation of the data.

For 17 out of 20 study hives, a model of collective foraging provided a better explanation of the data than one of individual foraging (Fig. 3a). In each of these cases, the collective model had a good fit (using a Kolmogorov-Smirnov statistic, see methods) to the empirical waggle run durations, and for 9 of these sites, the scout model was significantly different to the data (Fig. 3b). The sites shown in Figs 3c-d are representative examples showing the model fits, showing the overall quality of the model fit to the data.

```{r, fig.cap="Figure 3. A. Location of study hives in Southern England coloured by which model, collective (black circles) or individual (red triangles) provided the best explanation of the data as indicated by lowest AIC score. Sites STU and ZSL (C & D) are marked. Inset plot shows the location of the sites within the UK. B. Distribution of goodness of fit confidence values for each model fit to waggle run durations from each site. P value is derived from a bootstrapped two-sided KS test comparing the fitted model predictions to the empirical data. The red dashed line marks the significance threshold of 0.05. Values greater than 0.05 indicate no statistically significant difference is observed between the model and the data, indicating the model provides a good fit to the data. C. Model fits to waggle run durations collected from the ZSL hive along with histogram showing distribution. D. Model fits to waggle run durations collected from the STU hive along with histogram showing distribution. Model fits show the compliment cumulative frequency distribution, giving the probability of sampling a value greater than or equal to x."}
knitr::include_graphics("../results/figures/sites_model_plot.png")
```

Our results show that, whilst colony-level foraging mostly use a mixture of scouts and recruits, in some circumstances, colony foraging is best explained by a model of individual foraging. This suggests that in some environments, honey bees actively ignore the waggle dance and instead prioritise the use of private information. Previous empirical studies have shown that individual foragers flexibly alter their use of waggle dance information in response to unrewarding resources [@Gruter2011; @Price2019]. Evaluating how these individual decisions influence the collective, however, has historically been a challenge as it is effectively impossible to track an individuals foraging behaviour over a landscape. As our results show that these individual behaviours accurately describe colony foraging in three different sites, these findings support the idea that these individual behaviours can dominate colony foraging.

Our finding that foraging in some landscapes is more accurately described by ignoring the dance and focusing on private information supports the hypothesis that the dance is used flexibly in response to environmental factors. As the collective model quantifies waggle dance use, the fit of our model provides a tool to evaluate how environmental factors influence waggle dance use, and thus how a colony values social and private information in different environmental circumstances. This allows for comparing waggle dance use across environments.

We compared the estimates of the proportion of scout and recruits for colonies in different environments. Each site used in this study is situated in a landscape comprised of different proportions of land-use types. To evaluate how these different land-use types influenced waggle dance use, we first classified the different land-use types of the area surrounding each site to obtain a standardised land-use profile for the urban and agri-rural environments separately (see methods). As many of the resulting land-use types covaried with each other, we performed a Partial Least Squares (PLS) analysis to determine the principal components that represent combinations of land-use types which explained the most variation in the proportion of scout dances.

In the agri-rural environment one combination of land use types explained ~52% of the variation in the proportion of scouts (beta regression: $R^2 = 0.52$, $\phi$ = 3.7, p < 0.05, Fig. 4. A). This principal component correlates positively with arable land and negatively with non-agricultural unmanaged green space, built-up areas and water (Fig. 4. B). From the land-use maps coloured by each landscape dominating PC1 (Fig 4) we can see that arable land and built-up areas show the most coverage. As arable land increases whilst unmanged green space and built-up areas decrease, the proportion of individuals prioritising private information increases.

```{r, fig.cap="Figure 4. A. Estimated proportion of scouts for each site in the Agri-rural landscape against the first principle component derived from a Partial Least Squares analysis of land-use type. A beta regression shows the relationship (black line) between PC1 and the proportion of scouts, with 95% CI shown by the blue shaded area. B. Loadings for PC1 showing the correlations between PC1 and each land-use type (NAUMGS stands for Non-agricultural unmanaged green space). All points outside the shaded area significantly contribute to PC1. The colours of these land-use types are shown in the local environment of sites MEL, ROT and BEL (maps in bottom left) chosen as representative sites along the model gradient."}
knitr::include_graphics("../results/figures/lu_prop_scouts_agrirural.png")
```

A similar effect is also observed in the urban environment, where a single PC explained ~59% of the variance in the proportion of scouts (beta regression: $R^2$ = 0.59, $\phi$ = 6.2, p < 0.05, Fig. 5. A). This PC correlates positively with parks, allotments and cemeteries, whilst correlating negatively with sparse residential lands and railways (Fig. 5. B). Visualising how waggle dance use is influenced by changes in these land-use types shows that when the landscape is dominated by sparse residential land, as in site HHS, waggle dance use dominates the colonies foraging strategy. However, this reduces as sparse residential land is replaced by parks, allotments, and cemeteries (Fig. 5).

```{r, fig.cap="Figure 5. A. Estimated proportion of scouts for each site in the urban landscape against the first principle component derived from a Partial Least Squares analysis of land-use type. A beta regression shows the relationship (black line) between PC1 and the proportion of scouts, with 95% CI shown by the blue shaded area. B. Loadings for PC1 showing the correlations between PC1 and each land-use type. All points outside the shaded area significantly contribute to PC1. The colours of these land-use types are shown in the local environment of sites HHS, MPA and CAD (maps in bottom left) chosen as representative sites along the model gradient."}
knitr::include_graphics("../results/figures/lu_prop_scouts_urban.png")
```

In agri-rural environments, land-use types which typically associate with resource scarcity correlate positively with a higher proportion of scouts, whilst those which correlate with a higher resource abundance correlate with a lower proportion of scouts. Arable land is typically nutritionally poor for honey bees (ref). On the contrary, non-agricultural unmanaged green space is typically nutrient rich, as are built up areas in agri-rural environments, which often have gardens with ample flowers available for pollinators (ref). These findings indicate that in agri-rural environments, when resources are more abundant and thus easier to find, the collective foraging is used more, presumably to identify the best resources in the environment (ref). Conversely, where resources are more difficult to find, a decreased use of the waggle dance suggests the colony focuses more on individual exploration rather than in trying to locate on the best resources. A similar result is observed for urban environments, where the dependence appears to be driven by the presence of flower rich gardens in residential areas. However, an increase in the amount of parks, allotments and cemeteries decreases the use of the dance.

The fitting of our model to waggle dance observations provides a novel tool to quantify the relative importance of social and private information in honeybees. Further, by identifying the land-use combinations which most influence variation in the proportion of scouts, our analysis allows us to map the environment along the major axis of honeybee information use and visualise how land-use influences the use of the waggle dance. Our results within the agri-rural sites are in broad agreement with other studies which evaluated foraging performance with and without the dance [@Dornhaus2006; Price2019]. These results suggest that, even though collective foraging is not always beneficial, that in environments where high quality resources are present, but relatively scarce, collective foraging is being carried out by colonies, suggesting that under these conditions exchanging social information through the waggle dance confers benefits. These benefits of dancing appear to outweigh the costs associated with dancing and could explain why this remarkable behaviour is present in modern-day bees. However, the results for our urban sites do not produce a clear biological rational and present an all together more varied picture. More waggle dance data from urban colonies is therefore required to further probe how honeybees use the waggle dance in this heavily anthropogenic environment.

The analysis of waggle dance data can make an important contribution to our understanding of social information use and provides a methodology to further evaluate how honey bees use their unique dance language. Recently, technological advances have emerged which enable colony metrics to be collected faster, more accurately and over greater time spans than could be gathered by hand, allowing individuals to be tracked within colonies and theories of individual behaviour to be evaluated in more depth than could have been done previously [@Wild2021]. By piecing together the behavioural response of individuals and combining these with landscape analyses, we have found a mathematical description of colony foraging which extends our ability to predict individual behaviour beyond the colony walls and in to the environment. With the advances in the decoding of the waggle dance through automated methods [@Wario2017], we face the prospect of waggle dance data becoming “big data”. Our methodology thus provides a means of analysing such large data sets to inform the debate about the importance of social information, as well as providing useful colony metrics of foraging activity.
